\documentclass{svproc}

\usepackage{float}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{color}

\renewcommand{\floatpagefraction}{.95}
\renewcommand{\topfraction}{.95}

\definecolor{blue}{rgb}{0.13,0.13,1}
\definecolor{green}{rgb}{0,0.5,0}
\definecolor{red}{rgb}{0.9,0,0}
\definecolor{grey}{rgb}{0.46,0.45,0.48}

\lstset{language=Java,
  commentstyle=\color{green},
  keywordstyle=\color{blue},
  stringstyle=\color{red},
  basicstyle=\ttfamily}

\lstdefinelanguage{ddlog}{
  language=Java, % we base it on Java, just for comments
  morekeywords={input, output, typedef, relation, typedef, bool,
    string, bit, extern, function, var, for, match, skip, in,
    Aggregate, FlatMap},
  deletestring=[b]{'}
}

\newcommand{\LR}[1]{\textbf{\color{red}LR: #1}}

\author{
        Leonid Ryzhyk \and
        Mihai Budiu}
\institute{VMware Research}

\title{Differential Datalog}

\date{}

\begin{document}
\maketitle

\section{Abstract}

%In distributed systems, the control plane is responsible for
%reconfiguring the system in response to external events.  Due to
%performance and scalability requirements, it is impractical to
%recompute the entire configuration on every update.  Instead,
%developers employ \emph{incremental algorithms} that update their
%outputs in response to input changes.  Such algorithms are notoriously
%hard to develop, leading to complex and inflexible implementations.
%(This is the classic ``incremental view update''
%problem~\cite{Gupta-sigmod93}).


Many real-world applications of deductive databases require
incrementally updating output tables in response to changes to input
tables.
To address this need, we have developed Differential Datalog
(DDlog), a dialect of Datalog that automates incremental computation.
A DDlog programmer only has to write a specification for the
original (non-incremental) problem.  Given this description, the DDlog
compiler generates an efficient incremental implementation.  The DDlog
runtime automatically maintains indexes required to efficiently
compute output updates.

The DDlog language is targeted for system builders.  In consequence,
the language emphasizes usability, by providing a rich type system,
a powerful expression language, including string manipulation, arithmetics,
and control flow constructs.  Finally, DDlog offers an
alternative syntax for writing Datalog rules, which has a more
imperative flavor.

\input{intro}
\input{ddlog}
\input{system}
\input{applications}
\input{related}
\input{conclusions}

\bibliographystyle{abbrv}
\bibliography{top}

\end{document}
