\section{DDlog Implementation}\label{sec-system}

\subsection{Compiling DDlog to Differential Dataflow}

Figure~\ref{fig:compiler-flow} shows how DDlog programs are compiled
into native code.  The DDlog compiler is written in Haskell.  The
compiler performs parsing, type inference, validation, and several
optimization steps; it generates Rust code (as text files); the Rust
code is compiled and linked with the Differential Dataflow (DD)
library~\cite{differential-dataflow} to produce a static or dynamic
native code library.  DD is an independent open-source
project~\cite{differential-dataflow}, described in a series of
publications~\cite{timely-dataflow,differential-dataflow-paper} and
online documents~\cite{dd-mdbook,dd-reference}.  DD can be executed
using multiple cores or even multiple shared-nothing machines, but we
currently only support a single machine.

The output of the DDlog compiler, and the input to the DD library is a
dataflow graph, which may contain cycles (cycles are introduced by
recursion).  The edges of the dataflow graph are typed relations, and
the nodes are dataflow relational operators.  DD natively implements
the following operators: join, antijoin, distinct, union, aggregation,
filter, and flatmap.  Each operator has a highly optimized
implementation, incorporating temporal indexes that keep multiple
versions of each relation and allow efficient incremental evaluation.
The DD runtime library is responsible for executing the emitted
dataflow graph across many cores, in a transactional fashion.  The
DD runtime can be allocated a specific number of threads.

\subsection{Transactional API}

The interaction with a running DDlog program is done through a
transactional API.  At any time only one transaction can be
outstanding.  After \texttt{start}ing a transaction the users can
insert and delete any number of tuples from input relations.  When
attempting to \texttt{commit} a transaction all updates are applied
atomically and changes to all output relations are produced.  Users
can register an upcall to be notified of these changes.  These
notifications will happen on different threads, while the thread
executing the \texttt{commit} is blocked.  Figure~\ref{fig:javaapi}
shows an example Java program that interacts with a DDlog program.

\subsection{The command-line interface}

For every DDlog program the compiler generates an executable library,
a Rust program that allows invoking the DDlog program from Rust, and a
command-line interface program (CLI) that allows users to interact
with the DDlog program directly.  The CLI allows users to start
transactions, insert or delete tuples in input relations, commit
transactions, dump the contents of relations, and get statistics
about resource consumption.  The CLI is used for regression testing
and debugging.

\subsection{C and Java interfaces}

The compiled DDlog program exposes a transactional API to add and
delete records from input relations and notify the application about
changes in output relations at runtime.  The API is natively
implemented in Rust, with bindings available for other languages,
currently C and Java.  The Java API is program-independent, but the
native executable needs to be regenerated every time the DDlog program
is changed.  We illustrate the Java API to a simple DDlog with a
skeleton example in Figure~\ref{fig:javaapi}.

\begin{figure}[t]
  \footnotesize
  \begin{lstlisting}[language=ddlog]
// DDlog program    
input relation Parent(parent: string, child: string)
output relation Ancestor(ancestor: string, descendant: string)
Ancestor(ancestor, descendant) :- Parent(ancestor, descendant).
Ancestor(ancestor, descendant) :- Ancestor(ancestor0, ancestor1),
                                  Parent(ancestor1, descendant).
  \end{lstlisting}
  
  \begin{lstlisting}[language=Java]
// Java program interfacing with DDlog program    
static class Parent {
  String parent;
  String child; }
static class Ancestor {
  String ancestor;
  String descendant; }

/* Instantiate the DDlog program; add a record to the
 * Parent relation. */
static void main() {
  DDlogAPI api = new DDlogAPI(2 /* threads */,
               r -> onCommit(r) /* callback */);
  int parentRelation = api.getTable("Parent");
  Parent p = new Parent();
  p.parent = "Mike"; p.child  = "John";
  /* Create DDlog insert command. */
  DDlogCommand command = new DDlogCommand(
     DDlogCommand.Kind.Insert, parentRelation, p);
  int exitcode = api.start();  // start transaction
  DDlogCommand[] commands = new DDlogCommands[1];
  commands[0] = command;
  exitcode = api.applyUpdates(commands);
  /* commit transaction; causes the callback to be 
   * invoked for each new or deleted output record.
   * commit blocks until all upcalls have been invoked. */
  exitcode = api.commit();
  api.stop();
}

/* Callback invoked on commit once for each insertion or
 * deletion in an output relation. */
static void onCommit(DDlogCommand command) {
  int outputRelation = command.tableId;
  Ancestor value = command.getObject<Ancestor>();
  if (command.kind == DDlogCommand.Kind.Insert) 
    // ancestor inserted
  else 
    // ancestor deleted
}
\end{lstlisting}
\caption{Example program showing the Java API to pass and receive data
  from a DDlog program.\label{fig:javaapi}}
\end{figure}
