\documentclass{svproc}

\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{color}

\definecolor{blue}{rgb}{0.13,0.13,1}
\definecolor{green}{rgb}{0,0.5,0}
\definecolor{red}{rgb}{0.9,0,0}
\definecolor{grey}{rgb}{0.46,0.45,0.48}

\lstset{language=Java,
  commentstyle=\color{green},
  keywordstyle=\color{blue},
  stringstyle=\color{red},
  basicstyle=\ttfamily}

\lstdefinelanguage{ddlog}{
  language=Java, % we base it on Java, just for comments
  morekeywords={input, output, typedef, relation, typedef, bool,
    string, bit, extern, function, var, for, match, skip, in,
    Aggregate, FlatMap},
  deletestring=[b]{'}
}

\newcommand{\LR}[1]{\textbf{\color{red}LR: #1}}

\author{
        Leonid Ryzhyk \and
        Mihai Budiu}
\institute{VMware Research}

\title{Differential Datalog}

\date{}

\begin{document}
\maketitle

\section{Abstract}

In distributed systems, the control plane is responsible for
reconfiguring the system in response to external events.  Due to
performance and scalability requirements, it is impractical to
recompute the entire configuration on every update.  Instead,
developers employ \emph{incremental algorithms} that update their
outputs in response to input changes.  Such algorithms are notoriously
hard to develop, leading to complex and inflexible implementations.
(This is the classic ``incremental view update''
problem~\cite{Gupta-sigmod93}).

To address this complexity, we have developed Differential Datalog
(DDlog), a dialect of Datalog that automates incremental view updates.
A DDlog programmer only has to write an algorithm for solving the
original (non-incremental) problem.  Given this description, the DDlog
compiler generates an efficient incremental implementation.  The DDlog
runtime automatically maintains the indexes required for efficiently
computing view updates.

The DDlog language is targeted for system builders.  In consequence,
the language emphasizes usability, by providing a rich set of types,
an powerful expression language, including string manipulation, and
an alternative syntax for defining relations, which has a more
imperative flavor.

\input{intro}
\input{ddlog}
\input{system}
\input{applications}
\input{related}
\input{conclusions}

\bibliographystyle{abbrv}
\bibliography{top}

\end{document}
